\section{Introduction}

--- STILL DRAFTING ---

Computational analysis has become an integral part of modern-day plant pathology. High-quality genome sequences are being generated for multiple plant and pathogen species including banana and \ac{Fo}. The \ac{ncbi} genome database currently holds 17 genome entries for the Musaceae family and 736 genome entries for \ac{Fo}; 17 of which are from \ac{Focub}, including a high-quality draft genome for \ac{Focub4} \parencite{Warmington2019} and \ac{Focub1} \parencite{Asai2019} (\href{https://www.ncbi.nlm.nih.gov/data-hub/genome/}{https://www.ncbi.nlm.nih.gov/data-hub/genome/} Accessed: 17/01/2024). Many bioinformatic tools have been developed to analyse the increasing number of genome sequences. For instance, \textcite{Park2010} highlight integrated platforms supporting strain identification, phylogenetics, comparative genomics and knowledge sharing for \textit{Fusarium}. The use of genomics to identify candidate effectors in \ac{Fo} is becoming commonplace and is contributing to our understanding of \ac{Fo} classification, evolution, and cryptic host specificity. 

Generally, fungal effectors do not share sequence homology or conserved motifs, so computational prediction of fungal candidate effectors is challenging \parencite{Sperschneider2022, Todd2022}. Broadly, fungal effectors have been predicted using a set of common characteristics; the presence of a secretion signal, sequence length $\leq$ 300 amino acids, and cysteine richness \parencite{Sperschneider2015}. Other features sometimes used to filter protein sets include the lack of a transmembrane domain; increased expression during host interaction; restricted taxonomic distribution without (or with minimal) sequence resemblance to other organisms; and encoded by genes featuring extended intergenic regions or located in gene-sparse, repeat-rich chromosomes \parencite{Dalio2018, Todd2022}. However, as \textcite{LoPresti2015, Sperschneider2015} stress, not all secreted proteins with small size and high cysteine content necessarily serve as effectors. Conversely, fungal effectors are not universally small and cysteine-rich. Larger proteins can also function as effectors, which leads \textcite{LoPresti2015} to describe the 300 amino acid cutoff as arbitrary. Further, effector classification often relies on the absence of detectable orthologous proteins beyond the genus, but some effectors may show conservation or possess conserved functional domains \parencite{Jonge2010, Djamei2011, Mentlak2012}. Given these uncertainties in effector definition, \textcite{LoPresti2015} adopt a broad perspective, considering any secreted fungal protein as a potential effector. 

\textcite{Sperschneider2016} developed a \ac{ml} tool, EffectorP v1.0, for the identification of effectors in fungi in an attempt to combat the challenges of effector discovery explained by \textcite{Sperschneider2015, LoPresti2015}. 
EffectorP v1.0 is a Naïve Bayes \ac{ml} predictor, which was initially trained with 58 true fungal effectors from 16 species and achieved sensitivity and specificity of over 80\% \parencite{Sperschneider2016}. The negative dataset contained 14,143 proteins based on the total set of predicted secreted proteins from their 16 species (filtering the known effectors and homologs), and likely encompassed undiscovered effectors as well as non-effectors. Subsequently, EffectorP v2.0 was trained with 94 secreted true effectors from 23 species, utilising a refined negative dataset. This updated version successfully reduced effector candidates from fungal plant symbionts and saprophytes by 40\%, surpassing EffectorP v1.0 \parencite{Sperschneider2018}. The latest iteration, EffectorP (3.0), introduces a classification based on apoplastic and cytoplasmic localisation \parencite{Sperschneider2022}. EffectorP v2.0 has been employed by \textcite{Simbaqueba2020} in their analysis of \ac{Fo} f. sp. \textit{physalis}, identifying novel candidate effectors.

\Acp{mimp} provide an advantage when attempting to identify candidate effectors in \ac{Fo} genomes, as they are found upstream of all known effectors in \ac{Fol} \parencite{Schmidt2013} (see: section \ref{Chap1:fusariumEffectorome}). Indeed, \textcite{Dam2016} developed a computational pipeline (hereafter: FoEC) which identifies effectors using \acp{mimp}, to reveal effector profiles of cucurbit-infecting \ac{Fo} strains. The FoEC pipeline first identifies \acp{mimp} in a \ac{Fo} genome by searching for the \ac{mimp} \acp{tir} through regular expression. Once \acp{mimp} have been identified, FoEC expands 2500bp downstream of the \ac{mimp}, finding \acp{orf} using a custom Python script or using AUGUSTUS \parencite{Stanke2006} for gene prediction. Sequences are then filtered based on the presence or absence of a signal peptide, filtered for redundancy, and then searched against the input genome database using BLAST to create a presence/absence matrix. However, our analysis revealed that FoEC (v1) pipeline is unable to identify \acp{mimp} which have been soft-masked, a probable eventuality given that \acp{mimp} are repetitive elements. Removal of soft-masking to identify \acp{mimp} using FoEC (v1) will likely result in inaccurate gene predictions. Furthermore, FoEC (v1) only searches for effectors found downstream of a \ac{mimp}, but the study by \textcite{Schmidt2013} demonstrated that effectors may also be found upstream of a \ac{mimp}. 

FoEC has since been updated \parencite{FoEC2} (FoEC2)\footnote{Though it is not clear that the \ac{mimp} identification script has been adjusted to allow for soft-masking, it is of note that FoEC2 was developed and published following a discussion I had with the original FoEC authors.}, and uses a very similar pipeline structure. FoEC2 now also filters \acp{orf} based on cystine content and size (20aa $\leq$ size $\leq$ 600aa), clusters candidates using Diamond (v2.0.13) with Diamomnd BLASTP, MAFFT (v7.490) \parencite{Katoh2019} and uses HMMER \parencite{Eddy2011} to identify candidate effectors in the original genome database. FoEC2 now contains some of the filtering criteria which, as explained by \textcite{Sperschneider2015, LoPresti2015, Todd2022}, can miss candidates. 

Though we recognise that FoEC2 can be used as a tool for the identification of candidate effectors in \ac{Fo}, we developed a \ac{maei} pipeline\footnote{Development of the \ac{maei} pipeline was started before FoEC2 was published.} to find candidate effectors in \ac{Fo} genomes and contribute to our understanding of virulence within the \ac{FOSC}. Our pipeline considers the concerns of \textcite{LoPresti2015, Sperschneider2015, Todd2022}, instead identifying small ($\leq$450aa), secreted proteins associated with \acp{mimp} (upstream and downstream, 2500bp) in the \ac{Fo} genomes, submitting them to EffectorP v2.0 for candidate effector prediction. We applied the \ac{maei} pipeline to investigate candidate effector profiles in \ac{r1} and \ac{tr4} \ac{Focub} genomes, as well as \ac{r1} and \ac{r4} genome assemblies of \ac{Fola}. Candidate effectors identified using the \ac{maei} tool can be used as a target for molecular diagnostics, something which is currently being explored with \ac{r2}, \ac{r3}, and \ac{r4} isolates of \ac{Foa}. 

\newpage
\section{Materials and Methods}

\subsection{\textit{Fusarium} genome database generation}\label{chap3:fusariumdb}

\subsubsection{\textit{Fusarium} genome sequencing and assembly}

A database of both publicly available and new \textit{Fusarium} genome assemblies was produced for candidate effector identification and analysis (Table \ref{tab:GenomeDB})\footnote{The SY-2 isolate was included, but the S6, S16, and S32 genome assemblies were not used in this analysis as the raw read data were not available at the time.}. Collaborators at the \ac{niab} generated genome assemblies for \ac{Fo} \ac{fsp} \textit{lactucae} (\acs{Fola}) isolates AJ520, AJ718, AJ865, AJ516, AJ592 and AJ705; \textit{apii} (\acs{Foa}) isolates AJ498 and AJ720; \textit{coridandrii} (\acs{Foci}) isolate AJ615; \textit{matthiolae} (\acs{Foma}) isolate AJ260; \textit{narcissi} (\acs{Fona}) isolate FON63; and a \ac{Fo} isolate pathogenic towards rocket (\textit{Eruca vesicaria}), AJ174. These isolates were cultivated on \ac{pda} plates at 25°C for 4 days, and mycelial plugs were used to inoculate \ac{pdb} containing streptomycin. After 4 days of incubation, mycelium was harvested, lyophilised, and stored at -80°C. 

DNA extraction from 20mg of lyophilised mycelium was performed using a Macher\-ey-Nagel NucleoSpin Plant II kit (Fisher Scientific) with additional chloroform extraction after cell lysis. DNA samples were eluted in 30µl 10mM Tris HCl pH8 and analysed for quantity and purity using Nanodrop and Qubit (ratios 260/280 between 1.8-2.0, 260/230 between 2.0-2.2)  and for integrity using the Agilent TapeStation (TS 4150, Agilent Technologies). DNA samples (molecular weight >50kb) underwent Illumina PCR-free and \ac{ont} long-read sequencing. Illumina sequencing was conducted as described previously \parencite{Armitage2018}. \ac{ont} sequencing utilised the ligation sequencing kit (LSK108 for \ac{Foma} AJ260, \ac{Fona} FON63; or LSK110 all \ac{Fola}, \ac{Foa}, \ac{Foci} and \ac{Fo} rocket isolates) with flow cells FLO-MIN106 R9.4 (\ac{Foma} AJ260, \ac{Fona} FON63) or FLO-MIN106 R9.4.1 (\ac{Fola}, \ac{Foa}, \ac{Foci} and \ac{Fo} rocket isolates). 

Collaborators used \Ac{ont} long-read sequence data to construct \textit{de novo} genome assemblies. NanoPlot (v1.30.1) \parencite{DeCoster2018} was used for quality control of \ac{ont} data and Porechop (v0.2.4) (\href{https://github.com/rrwick/Porechop}{https://github.com/rrwic\-k/Porechop}), with default settings, was applied for adapter trimming. Filtlong (v0.2.1)  (\href{https://github.com/rrwick/Filtlong}{https://github.com/rrwick/Filtlong}) was used to remove reads shorter than 1 Kb or with a quality score less than Q9. Long-read data were assembled using NECAT (v0.0.1\_update20200803) \parencite{Chen2021} (genome size of 50 Mb) with parameters left as default. For long-read error correction, reads were aligned to the assemblies with Minimap2 (v2.17-r941) \parencite{Li2018} and used to inform one iteration of Racon v1.4.20 \parencite{Vaser2017}. This was followed by one iteration of Medaka (v1.5.0) (\href{https://github.com/nanoporetech/medaka}{https://github.com/nanoporetech/medaka}) using the r94\_min\_high\_g360 model. Illumina paired-end reads underwent quality control with FastQC (v0.11.9) (\href{https://www.bioinformatics.babraham.ac.uk/projects/fastqc/}{https://www.bioinformatics.babraham.ac.uk/projects/fastqc/}), followed by adapter and low-quality region trimming using Fastq-Mcf (v1.04) \parencite{Aronesty2013}. Alignment to long-read assemblies was conducted using Bowtie2 (v2.2.5) \parencite{Langmead2012} and SAMtools (v1.13) \Parencite{Danecek2021}. Three rounds of polishing with Pilon (v1.24) \parencite{Walker2014} corrected single base call errors and small insertions or deletions. Assembly statistics were computed using a custom Python script, and single copy ortholog analysis utilised \ac{busco} (v5.2.2) \parencite{Simao2015} with the hypocreales\_odb10 database.

\subsubsection{Publicly available \textit{Fusarium} genomes}

Two \ac{Fs} and 12 \ac{Focub} assemblies identified following GenBank genome search (\href{https://www.ncbi.nlm.nih.gov/data-hub/genome}{https://ww\-w.ncbi.nlm.nih.gov/data-hub/genome/}) were downloaded, as well as two \ac{Focub} genome assemblies (isolates 58 and 60) from the \ac{ngdc}, (\href{https://ngdc.cncb.ac.cn}{https:/\-/ngdc.cncb.ac.cn/}). Further, genome assemblies for the \ac{Foa} isolates AJ720 and AJ498 have already been published by \textcite{Henry2020} (AJ720 = GCA\_014843455.1,  AJ498 = GCA\_014843565.1) so were used in this analysis for comparison. \Ac{Foa} \ac{r3} and \ac{Foci} genome assemblies generated by \textcite{Henry2020} were also included. A representative assembly (\( \leq \) 50 contigs and \(\geq \) 90\% complete BUSCOs reported) was included for other publicly available \textit{Fusarium} species and \ac{fsp}. As \textcite{Schmidt2013} reported no \acp{mimp} in the \ac{Fg} PH-1 genome, a reference assembly for \ac{Fg} PH-1 (GCA\_000240135.3) was included as an outgroup for phylogenies and negative control for \ac{mimp}-associated effector analysis.

\input{Tables/GenomeDatabase}

\subsection{Phylogenetic analysis of \textit{Fusairum} isolates}

\Ac{tef} and \ac{rbp2} phylogenies were generated as previosuly described (section:~\ref{chap2:phylogeny}) for the \textit{Fusarium} assemblies included in the candidate effector analysis. Briefly, homologues of each \ac{tef} and \ac{rbp2} were identified in each assembly in our database  (Table \ref{tab:GenomeDB}) using BLASTN (1e-\textsuperscript{6} cut-off). The locations of hits with greater than 70\% identity and 90\% coverage were extracted using SAMtools (v1.15.1). \Ac{tef} and \ac{rbp2} sequences from each genome were concatenated into a single FASTA file for each barcode and aligned using MAFFT (v7.505) \parencite{Katoh2019}. IQ-TREE (v2.2.0.3) \parencite{Nguyen2015} was used to infer a maximum-likelihood phylogeny using the ultrafast bootstrap setting for 1000 bootstrap replicates and was visualised using iTOL \parencite{Letunic2021}. 



%\subsection{Identification of \textit{SIX} genes in \textit{Fusarium oxysporum} genome assemblies}

%As described previously (section: \ref{chap1:tnauSIXgenePhylo}), the complement of \ac{sixg} homologues was determined in all the \ac{Fo} genomes generated in this study, as well as in publicly available genomes (Table \ref{tab:GenomeDB}). \Ac{sixg} phylogenies were  constructed for the \acp{sixg} present in \ac{Fola} \ac{r1} (SIX9, SIX14) and \ac{Fola} \ac{r4} (SIX8, SIX9, SIX14). 

\subsection{\textit{Mimp}-associated effector identification (Maei) pipeline}

\subsubsection{\Ac{mimp} identification}

Two methods of \textit{mimp} searching were developed. The first identifies \acp{mimp} using a custom Python (v3.8.8) script \parencite{Python} which uses the Biopython (v1.78) package \parencite{biopython}. As performed in \textcite{Schmidt2013, Dam2016, Armitage2018} \acp{mimp} were identified in each \textit{Fusarium} genome using regular expression, whereby the \textit{mimp1-4} consensus \ac{tir} sequences \parencite{Bergemann2008, Schmidt2013}, "CAGTGGG..GCAA[TA]AA" and "TT[TA]TTGC..CCCACTG", were used as a search pattern. Where sequences matching this pattern occured within 400 nucleotides of each other a \ac{mimp} was recorded (script available via \href{https://github.com/JamiePike/Maei/blob/main/bin/Mimp_finditer.py}{GitHub}). The second method employs a \acf{hmm} which was developed using the \ac{hmm} tool HMMER (v3.3.1) \parencite{Eddy2011}. Publicly available \ac{mimp} sequences (Appendix \ref{apx:RefMimps}) and \ac{mimp} sequences identified using the regular expression method were used to build a \ac{mimp} profile-\ac{hmm}. This profile-\ac{hmm} was used as the input for an NHMMER (v3.3.1) \parencite{Eddy2011} search of each genome (full command-line arguments available via GitHub: \href{https://github.com/JamiePike/Maei/blob/main/dev/notes/Mimp_Searching_Methods.md}{https://github.com/JamiePike/Maei/dev/notes/Mimp\_Searching\_Methods.md}).

\subsubsection{Sequence extraction}

Using \acp{mimp} identified by both \ac{mimp} searching methods, sequences 2.5kb upstream and downstream of each \ac{mimp} were extracted, merged (to create a non-redundant set), and stored in GFF3 format using the BEDtools (v2.25.0) suite \parencite{Quinlan2010} (full command-line arguments available on GitHub: \href{https://github.com/JamiePike/Maei/blob/main/bin/Maei_v5.sh}{https://github.com/JamiePike/Maei/bl\-ob/main/bin/Maei\_v5.sh}). Using the GFF3 locations, two FASTA files were generated i) "\ac{mimp}-region" FASTA where all nucleotides not within a 5kb window (2.5kb upstream + 2.5kb downstream) of a \ac{mimp} were hard-masked, ii) "AUGUSTUS-region" where all nucleotides outside of a 45kb window (22.5kb upstream + 22.5kb downstream) of a \ac{mimp} were hard-masked. The "AUGUSTS-region" FASTA file was generated with a 45kb window to prevent truncation of gene models. 

\Acp{orf} withing the "\ac{mimp}-regions" were identified using the EMBOSS (6.6.0.0) tool, getorf (https://www.bioinformatics.nl/cgi-bin/emboss/getorf). The "AUGUSTUS-region" FASTA was subjected to gene prediction using AUGUSTUS (3.3.3) \parencite{Stanke2006} with the “Fusarium” option enabled. The AUGUSTUS (v3.3.3) gene models and "\ac{mimp}-region" GFF3 files were intersected and merged using the BEDtools (v2.25.0) suite \parencite{Quinlan2010} to ensure that gene models which may be partially within the 5kb "\ac{mimp}-region" window are not truncated, but that gene models outside of the 5kb "\ac{mimp}-region" window but within the 45kb "AUGUSTUS-region" window are not included. The intersected AUGUSTUS (v3.3.3) gene models and "\ac{mimp}-region" GFF3 file was used as input for AGAT (v1.0.0) \parencite{DainatN.D.}, which was used extract AUGUSTUS (v.3.3.3) \ac{mimp}-associated gene models in FASTA file format. The \ac{mimp}-associated gene model FASTA file and getORF FASTA were merged to generate a \ac{mimp}-associated gene model and \ac{orf} FASTA file. The output from getORF was also converted to GFF3 format using a custom Python script (available via \href{https://github.com/JamiePike/Maei/blob/main/bin/getORF2bed.py}{ GitHub https://github.com/JamiePike/Maei/blob/main/bi\-n/getORF2bed.py}
). These GFFS were then merged to generate a \ac{mimp}-associated gene models and \acp{orf} GFF3 and FASTA file for each genome using BEDtools (v2.25.0) suite.

\subsubsection{Candidate effector filtering and prediction}

Sequences from the \ac{mimp}-associated gene model and ORF FASTA were then filtered based on size using SAMtools (v1.6) and the BEDtools (v2.25.0) suite, with sequences >30aa and <450aa submitted to SignalP (v4.1) \parencite{Petersen2011}. Sequences that were predicted to contain a signal peptide were passed to EffectorP (v2.0.1) \parencite{Sperschneider2018} for candidate effector prediction. This output is was to generate a genome-specific candidate effector sets.

\subsubsection{Candidate effector clustering and distribution}

The candidate effector sets from each genome were then combined and clustered using CD-HIT (v4.8.1) \parencite{Fu2012} at 80\% identity, to generate candidate effector clusters. Data from CD-HIT (v4.8.1) was also passed to a custom Python script, which was used to generate an overview table and a distribution matrix (available via GitHub: \href{https://github.com/JamiePike/Maei/blob/main/bin/Processingcdhit.py}{https://github.com/JamiePike/Maei/blob/main/bin/Processingcdhit.py}).

To identify candidates which may be shared across isolates but not associated with a \ac{mimp} in all isolates, the combined candidate effector sets were then searched against the \textit{Fusarium} genomes using TBLASTN, with a cut-off 1e-6 and a percentage identity and coverage threshold of 60\%. A data matrix was generated using the TBLASTN hit data and a heatmap was generated using the R package Pheatmap. Sequences within the threshold were extracted using  SAMtools (v1.6) and the BEDtools (v2.25.0) suite and were subjected to filtering using SignalP (4.1), default settings and EffectorP (2.0.1) to remove unlikely candidates. 

A Mann-Whitney U test was used to assess the statistical significance of candidate effector content between genomes, and to determine if there is a correlation between genome size and candidate effector content, Spearman’s correlation was performed in R (v3.6.3) \parencite{R}.

\subsection{Expression and putative function of candidate effectors}

\subsubsection{\Acf{Fola} inoculation, RNA extraction and RNAseq analysis}

Collaborators at \ac{niab} inoculated lettuce seedlings with \ac{Fola} and non-pathogenic \ac{Fo} isolates and conducted RNA extraction and analysis. Their method was adapted from \textcite{Taylor2016}. Two variants of Fola4 (AJ516, AJ705) and Fola1 AJ520 were used, along with controls: Foma AJ260 (pathogenic on column stocks) and F. oxysporum Fo47 (non-pathogenic isolate) (isolates indicated in table \ref{tab:GenomeDB}). A non-inoculated control treatment was also established. Autoclaved ATS medium (1 M KNO\textsubscript{3}, 1 M KPO\textsubscript{4}, 1 M MgSO\textsubscript{4}, 1 M Ca(NO\textsubscript{3})\textsuperscript{2}, 20 mM Fe-EDTA, 70 mM H\textsubscript{3}BO\textsubscript{3}, 14 mM MnCl\textsubscript{2}, 0.5 mM CuSO\textsubscript{4}, 1 mM ZnSO\textsubscript{4}, 0.2 mM Na\textsubscript{2}MoO\textsubscript{4}, 10 mM NaCl, 0.01 mM CoCl\textsubscript{2}, 0.45\% Gelrite) (Duchefa Biochemie, Haarlem, The Netherlands) in square petri dishes (12 x 12 x 1.7 cm, Greiner Bio-One, UK) was used for growth. \ac{Fola}-susceptible lettuce seeds (cv. Kordaat) were surface-sterilised (10\% bleach water (v/v) solution for 5 min and then rinsed three times with \ac{sdw}) and placed on agar plates. Stacks of plates were chilled for 4 days, followed by incubation at 15°C for 8 days and then 25°C for six days. Conidial suspensions of each two-week old \ac{Fo} culture (grown on \ac{pda} at 25°C) was prepared, adjusted to 1 x 10\textsuperscript{6} spores mL\textsubscript{-1} in \ac{sdw} with the addition of 200 \(\mu\)L of Tween20 L\textsubscript{-1}, and 1.5mL pipetted directly onto lettuce roots (seven agar plates per isolate). Non-inoculated controls received \ac{sdw} + Tween. RNAseq samples were obtained from lettuce roots at 96 \ac{hpi} (four agar plates per isolate) and pooled in to one sample of 12 plants per plate, corresponding to the peak expression of \textit{SIX8}, \textit{SIX9}, and \textit{SIX14} (data not shown). Lettuce roots were rinsed in \ac{sdw}, flash frozen in liquid nitrogen and stored at -80°C before extraction. Disease assessment was conducted two \ac{wpi} (using three remaining plates per isolate).

In order to identify unregulated genes \textit{in planta}, \ac{Fo} isolates were cultured \textit{in vitro}. Spore suspensions (500 \(\mu\)L 1 x 10\textsuperscript{6} spores mL\textsubscript{-1}) were pipetted onto \ac{pda} plates (prepared as above) with autoclaved cellulose discs and incubated for 96 hours at 25°C. Mycelium was harvested by scraping off the layer growing on the cellulose surface followed by flash freezing. 

Lettuce roots were ground to a fine powder using a pestle and mortar filled with liquid nitrogen and approximately 100 mg of tissue transferred to a 2 mL tube. RNA was extracted using Trizol® reagent (Thermo Fisher Scientific). Extracted RNA was precipitated using \(\mu\)L of lithium chloride to 100 \(\mu\)L of RNA (250 μL LiCl2 + 650 μL DEPC treated water). DNase1 (Sigma-Aldrich) was used to remove DNA. To check degredation, RNA samples were visualised on a 2\% agarose gel (containing GelRedTM at 2 μL per 100 mL of gel) with Orange G (Sigma-Aldrich) loading dye. RNA samples were sent to  Novogene conducted polyA-enrichment, followed by Illumina PE150 sequencing at a depth of 23G raw data per sample for \textit{in planta} samples and 9G per sample for \textit{in vitro} mycelial RNA samples.

Transcripts from both \ac{Fo} isolates and lettuce RNA-Seq reads had  adaptees trimmed using Fastq-Mcf (v1.04) \parencite{Aronesty2013}. To distinguish \ac{Fo} reads from lettuce, the lettuce genome (\textit{Lactuca sativa}, \ac{ncbi} accession: GCF\_002870075.3) was obtained from \ac{ncbi}. Pre-processed reads were then aligned to respective F. oxysporum assemblies using \ac{star} (v2.7.10) \parencite{Dobin2013}, incorporating the flag --outReadsUnmapped to generate a file containing non-mapping reads. The non-mapping putative \ac{Fo} RNA-Seq reads were pseudo-aligned to the respective reference genome and quantified using Kallisto (v0.48.0) \parencite{Bray2016}. Subsequently, R (v4.1.3) \parencite{R} and the DESeq2 package (v1.34.0) \parencite{Love2014} were utilised for conducting differential gene expression analysis. The contrast function was applied to produce a list of differentially expressed genes.

\subsubsection{\textcolor{red}{Functional predictions}}

\subsection{Data and software availability}

The complete computational pipelines, along with command-line arguments and bash, R, and Python scripts employed for the candidate effector identification and analyses, can be accessed in the Maei GitHub Repository (\href{https://github.com/JamiePike/Maei}{https://github.com/JamiePike/Maei}). Previous versions and supporting development documentation is available in Markdown format for \ac{maei} \href{https://github.com/JamiePike/Maei/tree/main/dev}{pipeline development}.

\newpage
\section{Results}


% Plan for Figures:

% - Table of all isolates included

% - TEF phylogeny of isolates included in the assessments

% - Something on mimp ID? 

% - Ability of versions to predict effectors Table - Include FoEC.py first results?

% - Stats bar charts of assembly size and effector pred. 

% - Major heatmap of effectors 

% - Phylogeny and heatmap ggtreeExtra? 

% - Minor heatmap of race specific effectors (lactucae and banana)

% - Comparison to pred effectors from TNAU set (TBLASTN)

% - Pred functions of race specific effectors?

% - RNA seq of lactucae 

% - Lactucae SIX gene trees??? 

% - Apii diagnostics 


\newpage
\section{Discussion}

\subsection{}{Segmental duplication and the evolution of effectors}

van Westerhoven \et (2023) investigated the effects of segmental duplications on effector repertoires.  Though previously reported by van Dam \et (2016), the authors provide a more comprehensive effector gene repertoire for \ac{Focub} and demonstrate that effector repertoires were variable among the strains. The study found that 336 out of 669 predicted effector genes in \ac{Focub4} strain II5 evolved via segmental duplications. Notably, the effector SIX1, which is essential for full virulence to banana, was found to be part of a segmental duplication. This supports van Westerhoven \et (2023) argument and suggests that segmental duplications are involved in the evolution and diversification of effector genes in \ac{Focub}, which in turn influence the pathogenicity and virulence of \textit{Fusarium} strains.

Though we know that cytoplasmic fungal effectors are secreted into plant host cells, the mechanism by which fungal pathogens deliver effector proteins into plant cells remains poorly understood. Recently, \textcite{Oliveira-Garcia2023, Wang2023} have provided evidence for the uptake of cytoplasmic effectors from \textit{Magnaporthe oryzae} and \textit{Phytophthora infestans} into host \textit{Oryza sativa} or \textit{Nicotiana benthamiana} cells, respectively, through clathrin-mediated endocytosis (CME).  

\textcolor{red}{[ALSO, IF YOU USE EFFECTORS AS A TARGET, THEY ARE PRONE TO HIGH AMOUNTS OF EVOLUTIONARY PRESSURE, AND MAY MUTATE RENDERDING DIAGNOSTICS USELESS - YOU THEREFORE NEED A MEANS OF CONSISTENTLY INDENTIFYING NEW TARGETS...]}