
First, I would like to thank my supervisor, Prof. Murray Grant. Murray has challenged me to become more independent, thoughtful, and confident as a researcher, and I am grateful for the support, guidance, time and wisdom he has provided. Second, I would like to thank Dr David Studholme for his valuable supervision and feedback, particularly in the race to finish this thesis. Next, I would like to thank Prof. John Clarkson and Dr Helen Bates. The opportunity to trial my pipeline on their \textit{Fusarium} genomes was invaluable, and the insights they provided were extremely useful. I would have a substantially smaller thesis without them. 

My thanks to Dr Laura Baxter for her advice, ideas, support and expertise. For every panicked error message I sent, Laura had sage advice. A huge thank you to Dr John Sidda. Not only must I thank John for the help he provided with all things metabolomic, but also for welcoming me to the group, his friendship, and his WhatsApp puns. My thanks to Dr Lijiang Song, Dr Zac Newland-Smith, and Nadine Aschuer for their help in processing many metabolomics samples. I also thank Dr Raveendran Muthurajan at TNAU. Of course, a big thank you to all the members of C46 and the Grant group, past and present. 

This PhD would not have been possible without the funding provided by the Waitrose CTP, I am grateful to the CTP for the financial support and the wider Waitrose CTP training opportunities and events. 

Beyond the PhD, I would like to thank (Dr) Annabelle de Vries and (Dr) Sadik Muzemil for their help, friendship, and opportunities to chat, brainstorm, and panic in B1.34. On the subject of helpful friends, I must thank Hannah and Alissa for the many, many hours they spent proofreading this thesis, as well as their encouragement and, occasionally welcome, cat-related distractions. I would like to thank Tash, Ciaran, Rosie, Harry, and Delenn, as well as Dr James Wagstaffe, Dr Rachel Kaleta, and Ms Melanie Webb, Dr Kevin King, Dr Jon West, and Ms Gail Canning for their horticultural and scientific enthusiasm. My thanks to Justine, Emma, Tommy, Nan and everyone Pike, Warren,  Fleckney, or otherwise, who has encouraged and supported me.  

Finally, a special thanks to my partner, Mark, whose unwavering love, support, and understanding has been invaluable. Additionally, I am deeply grateful to my parents for their endless love, encouragement, and the sacrifices they have made to support my academic pursuits. Their guidance, belief in me, and unwavering support have been fundamental to my success, and I am profoundly grateful for everything they have done. 




