This thesis is submitted to the University of Warwick in support of my application for the degree of Doctor of Philosophy. It has been composed by myself and has not been submitted in any previous application for any degree. The work presented (including data generated and data analysis) was carried out by the author except in the cases outlined below. 
\noindent
\subsubsection{Data provided and/or analysis carried out by collaborators.}
\begin{itemize}
    \item Virulence assays and generation of genomic read data for \textit{Fusarium} isolates, S6, S16, S32, and SY-2. Data was supplied by Dr Raveendran Muthurajan and collaborators at \acl{tnau}. 
    \item Genome assemblies for some \acl{Fo} isolates (denoted in tablem\ref{tab:GenomeDB}) were supplied by Dr Helen Bates and collaborators at \acl{niab} and Prof. John Clarkson and group at the University of Warwick. 
    \item RNAseq data and supporting analysis for \acl{Fo} isolates (AJ516, AJ520, AJ705, AJ260, and Fo47) were supplied by Dr Helen Bates and collaborators at \acl{niab} and Prof. John Clarkson and group at the University of Warwick. 
    \item The \acl{tef} database used for phylogenetic analysis was compiled by Chloe Harper, as part of her undergraduate research project (supervised by Prof. John Clarkson, University of Warwick).
    \item Scripts used for IPO and XCMS analysis were supplied by Dr John Sidda, University of Oxford.
\end{itemize} 

\subsubsection{Publications including submitted papers.}
\begin{itemize}
    \item Parts of the analysis conducted in Chapter 2 have been submitted for publication, entitled: \textit{Comparative genomics and transcriptomics reveal differences in effector complement and expression between races of Fusarium oxysporum f.sp. lactucae}. 
\end{itemize}
